Distinguishing among coalescent models using two-site allele frequency spectra

### 2018-08-07 manuscript/manuscript.tex
Finish results. Begin methods.

After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtal2015}.
We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{table:DPGP_params} [Make table].
We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}).

@@

\begin{itemize}
    \item Expected and observed SFS match (because we fit the model to that).
    \item Both show excess of low-frequency SNPs.
    \item At intermediate distance, can see that the $\fpmi$ is strongly positive in the data, but negative in the model.
    \item This pattern extends to other genetic distances.
    \item Correlations in the data decay at long distances. This is like the sweeps simulations, but we don't want to draw too strong a conclusion.
\end{itemize}

@@

\section*{Methods}
\begin{itemize}
    \item Code to compute moments (currently piecewise-constant and exp. growth, but could be extended easily)
    \item Simulation code extending msprime
    \item Forward-time simulation with SLiM
    \item Drosophila
    \begin{itemize}
        \item data pre-processing: throw out inversions, downsampling to 100 chromosome, 4-D sites, filter for coverage
        \item picking the central regions
        \item fitting demographic model with fastNeutrino: fit 3-epoch piecewise constant to SFS. Report model parameters
        \item simulating demographic model
        \item computing statistics
    \end{itemize}
\end{itemize}

@@

@@

% Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).

>>>>

After filtering for coverage, removing chromosome arms with known inversions, downsampling to $n=100$ samples per autosomal chromosome arm, and identifying 4-fold degenerate sites, we selected the central region of each chromosome that have consistent high diversity (Methods).
Because the average pairwise diversity varies between arms---possibly reflecting selection or different sets of segregating inversions---we performed all subsequent calculations on each arm independently.
We fit a demographic model to the site frequency spectra of these central regions using \texttt{fastNeutrino} \cite{BhaskarEtal2015}.
We fit a 3-epoch piecewise-constant model, with four free parameters: two changepoints and two population size ratios.
We report our fitted parameters and those of a similar published model \autocite{RagsdaleGutenkunst2017} in Table~\ref{tab:DPGP_params} [Make table].
We then simulated under our fitted model using \texttt{msprime} and computed the expected and observed SFS, $\fpmi$, and $\hilopmi$ (\fig{fig:dpgp3}).

@@

The first row of \fig{fig:dpgp3} shows that the expected SFS under the fit demographic models agree with the observed SFS, demonstrating that a time-varying $N(t)$ can explain this aspect of the data well.
In contrast, the second row shows the expected and observed weighted $\fpmi$ for averaged over distances less than $15 d_c$ apart, where $d_c = 2/ (\pi \mu)$ is the distance scale corresponding to $r\E{T_2} = 1$.
Here, the data shows strong positive associations between nearby alleles at different frequencies, while the model of population growth predicts weak negative associations except just off of the diagonal.
This pattern extends across a range of genomic distances, $d/d_c \in (10^{-1}, 10^2)$ (\fig{fig:dpgp3}, third row).
As a result, we may conclude that the data is not well explained by the Kingman coalescent with population growth.

Note that the $\hilopmi$ decays toward zero at large distances.
This matches the expectation from simulations with selective sweeps rather than the beta coalescent.
However, we caution against concluding that sweeps are necessarily responsible for the deviations from the Kingman expectation.

@@

\section*{Methods}

\subsection*{Computing branch-length moments}

We implemented numerical computations of the moments of the branch lengths $\E{\tau_i}$ and $\E{\tau_{ij}}$ using the \texttt{python} numerical package \texttt{numpy} \autocite{numpy}.
For the Kingman coalescent with time-varying coalescent rate, we used equations~(1)-(12) of \cite{ZivkovicWiehe2008}.
To calculate the second moments of the coalescent time distribution for exponentially growing populations, we used Gaussian quadrature.
For the beta coalescent, we implemented the recursion described by \cite{BirknerEtAl2013}.

Functions for compute the branch-length moments are available [GITHUB].
The Kingman coalescent code currently can compute moments for exponentially growing and two-epoch piecewise constant models, but could be extended to allow for other models.
The formulas of \cite{ZivkovicWiehe2008} exhibit numerical instability related to the instability of \cite{GriffithsTavare1994}.
They are therefore only practical for sample sizes up to $n \approx 40$.
The recursion of \cite{BirknerEtAl2013} is $\mathcal{O}(n^4)$ [CHECK THIS] and is also only practical for samples up to $n \approx 50$.

\subsection*{Coalescent simulations}

Our beta coalescent simulation code is based on modifications to the \texttt{msprime} package made by Joe Zhu [link to github] to allow for multiple mergers.
This code, together with \texttt{python} wrapper scripts and utility functions to run simulations and calculate the 2-SFS and $\fpmi$ from the \texttt{msprime} output is available [GITHUB].

[Possibly move some of the description here]

\subsection*{Forward-time simulations of selective sweeps}

We simulated a model of recurring selective sweeps using the software \texttt{SLiM} \autocite{MesserEtAl201?}.
In all of these simulations, we simulated a population of 500 diploids for $10^4$ generations.
Each haploid genome consisted of a single genomic element $L=10^8$ basepairs long with recombination rate per basepair $r = 10^{-8}$ and overall mutation rate $\mu = 10^{-7}$.
We simulated two types of mutations: neutral mutations and beneficial mutations with additive effects and selection coefficient $s=0.1$.
With these parameters, $2Ns = 100$, so beneficial mutations are strongly selected and will sweep in $T_{\text{sweep}}\sim s^{-1} \log Ns \approx 50$ generations.
Such sweeps will effect a region of the chromosome $d_{\text{sweep}} \sim (r T_{\text{sweep}})^{-1} \approx 2 \times 10^6$ basepairs long.
Thus $d_{\text{sweep}} \ll L$, which will minimize edge effects of having a finite chromosome.

In order to vary the effects of sweeps on neutral diversity, we varied the fraction of mutations that are beneficial $f_{\text{sel}}$ over several orders of magnitude: $f_{\text{sel}} \in \{10^{-6}, 10^{-5}, 10^{-4}, 10^{-3}\}$.
For each $f_{\text{sel}}$, we ran 100 independent replicate simulations and computed $\eta_i$ and $\eta_{ij}$ of neutral mutations averaged over all replicates.

\texttt{SLiM} parameter files, \texttt{python} wrapper scripts for parsing output, and \texttt{snakemake} files for running simulations are available [GITHUB].

\subsection*{Analysis of \textit{D. melanogaster} data}

The DPGP3 data set consists of haploid consensus sequences from $\sim 200$ files, obtained via the haploid embryo method of \cite{LangleyEtal2011}.
We obtained the DPGP3 consensus sequence files version 1.1 from \url{www.johnpool.net/genomes.html}.
These files contain the sequence alignments of all flies in the sample on all chromosome arms.
We also downloaded the Nov. 3, 2016 spreadsheet of inversions available at the link above.
For each chromosome arm, we excluded any samples with an inversion in that arm and then down-sampled to $n=100$ by taking the first 100 remaining samples in alphanumeric order of sample name.
(This means that the data for each chromosome arm is from a slightly different subset of the individuals.)

\begin{itemize}
    \item data pre-processing: throw out inversions, downsampling to 100 chromosome, 4-D sites, filter for coverage
    \item picking the central regions
    \item fitting demographic model with fastNeutrino: fit 3-epoch piecewise constant to SFS. Report model parameters
    \item simulating demographic model
    \item computing statistics
\end{itemize}

@@

\section*{Tables}

\begin{table}[h!]
  \begin{center}
    \caption{[Table of parameters underlying Figure 2]}
    \label{tab:nonrec_params}
    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
      % $\alpha$ & $\beta$ & $\gamma$ \\
      % \hline
      % 1 & 1110.1 & a\\
      % 2 & 10.1 & b\\
      % 3 & 23.113231 & c\\
    \end{tabular}
  \end{center}
\end{table}

\begin{table}[h!]
  \begin{center}
    \caption{[Table of parameters underlying Figure 4]}
    \label{tab:n4_params}
    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
      % $\alpha$ & $\beta$ & $\gamma$ \\
      % \hline
      % 1 & 1110.1 & a\\
      % 2 & 10.1 & b\\
      % 3 & 23.113231 & c\\
    \end{tabular}
  \end{center}
\end{table}

\begin{table}[h!]
  \begin{center}
    \caption{[Table of parameters underlying Figure 6]}
    \label{tab:rec_params}
    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
      % $\alpha$ & $\beta$ & $\gamma$ \\
      % \hline
      % 1 & 1110.1 & a\\
      % 2 & 10.1 & b\\
      % 3 & 23.113231 & c\\
    \end{tabular}
  \end{center}
\end{table}

\begin{table}[h!]
  \begin{center}
    \caption{[DPGP3 model fit]}
    \label{tab:DPGP_params}
    \begin{tabular}{l|c|r} % <-- Alignments: 1st column left, 2nd middle and 3rd right, with vertical lines in between
      % \textbf{Value 1} & \textbf{Value 2} & \textbf{Value 3}\\
      % $\alpha$ & $\beta$ & $\gamma$ \\
      % \hline
      % 1 & 1110.1 & a\\
      % 2 & 10.1 & b\\
      % 3 & 23.113231 & c\\
    \end{tabular}
  \end{center}
\end{table}

@@

TODO:
- Abstract
- Drosophila methods
- Discussion
- Parameter tables:
    -
- bibtex file
- SI figures:
    - Drosophila diversity
    - Drosophila # called sites
- Code
% Tests of the standard Kingman coalescent based on the SFS originated with \cite{Tajima1989}, \cite{FuLi1993}, and \cite{SimonsenEtAl1995} (see \cite{Achaz2009} for a unifying framework).
